---
title: Demo
subtitle: Reproducible and Dynamic Reporting with R and Quarto
author:
  - name: 'Gorka Fraga González'
    orcid: 0000-0002-1857-8607

  - name: 'Fabio Molo'
    orcid: 0000-0001-9320-5854

  - name: 'Samuel Pawel'
    orcid: 0000-0003-2779-320X
format:
  html:
    toc-depth: 4
    df-print: paged
  pdf: default
execute: 
  eval: true
bibliography: references.bib
---

This demo requires R and Quarto Visit the [instructions](../instructions/index.qmd) page to get ready.

# Part I - Create your first report

## 1. Install necessary R packages

Run the following command in the R console. Installation may take a few minutes.

```r
install.packages(c("dplyr", "ggplot2", "kableExtra"))
```

## 2. Create a new Quarto document

### Create a new .qmd file

-   In the Rstudio menu, go to `File > New File > Quarto document`
-   Use the default settings and enter a filename
-   **Save** the file with a `.qmd` filename extension  in the project home directory (visible on the bottom right pane)
-   Switch between the `visual` and `source` **tabs** in RStudio
-   The `source` tab displays the raw code, while the `visual` tab shows a live preview of the formatted document. This makes it easier to see how your report will look when rendered

### Edit the YAML header

Edit the YAML[^1] header delimited by (`---`). Add information about `author`, `affiliation` and `date` (use `last-modified` to always display the latest date) and define the output format as `pdf` or `html`. Be careful with *indentation*, YAML is sensitive to spacing! The header can be kept simple or it can contain a lot of information. An example:


[^1]: YAML is an acronym for "Yet Another Markup Language"

``` yaml
---
title: "Title of the report"
subtitle: "Subtitle"
date: last-modified # display the date of the last edit
author: 
  - name: "Author 1"
    orcid: 1234-1234-1234-12345
    affiliation:
      - name: "University of Zurich"
        city: "Zurich"
  - name: "Author 2"
    orcid: 5678-5678-5678-5678
    affiliation:
      - name: "University of Bern"
        city: "Bern"
format:
  html:
    toc: true # add a table of contents
    toc-depth: 4 # show level 1-4 subheadings in the table of contents
  pdf: default
---
```

## 3. Create the report content

The two most important elements of a dynamic report are text and code.

### Text

Quarto uses Markdown syntax to structure and format text. For example,

``` markdown
# This is a title

Two stars will render a word **bold**. 

We can also create lists:

- Element 1
- Element 2

A useful link is this [Markdown guide](https://quarto.org/docs/authoring/markdown-basics.html)

> An impressive quote
```

produces

<h2>This is a title</h2>  <!-- HACK to not have the title in the TOC -->

Two stars will render a word **bold**.

We can also create lists:

-   Element 1
-   Element 2

A useful link is this [Markdown guide](https://quarto.org/docs/authoring/markdown-basics.html)

> An impressive quote


This type of text editing is different from WYSIWYG (*What you see is what you get*) editors, such as Microsoft Word. However, the `visual` mode in RStudio provides a WYSIWYG-like interface and can be useful if you do not know certain Markdown syntax.

### Code chunks

[Code chunks](https://quarto.org/docs/computations/r.html) contain executable code and let you embed live results directly into your text, for example, data summaries, plots, or tables. Code chunks are automatically re-run when the document is rendered. If any code chunk fails, the entire document may fail to render. This encourages writing reproducible, error-free code.

::: {#tip-chunks .callout-tip collapse="false"}
#### Code execution options

The code chunks are delimited by ```` ```{r} ```` (you can specify other languages like Python, Julia, ObservableJS). For each code you can add *execution options* to the code chunks with `#|`, for example `#| echo: false` will not display the code. In Rstudio you can see the options available if you type `#|` and then press the TAB key on your keyboard . These options can be set globally for the entire document on the YAML header (under `execute:`).
:::

#### Load libraries

We first load all R packages that we will use in our analysis. Copy the code below inside a "code chunk".

```{r}
#| warning: false
#| output: false
library("dplyr") # data wrangling
library("ggplot2") # plotting
library("kableExtra") # tables
```

#### Read the data

For this example we will use the data set "Vornamen neugeborener Mädchen und Knaben mit Wohnsitz in der Stadt Zürich" (<https://data.stadt-zuerich.ch/dataset/bev_vornamen_baby_od3700>). This data set contains the names of newborns in the city of Zurich since the year 1993. The data was kindly made available to the public domain via a CCZero license by Open Data Zürich. We will now import the data and rename the variables into English.

```{r}
url <- "https://data.stadt-zuerich.ch/dataset/bev_vornamen_baby_od3700/download/BEV370OD3700.csv"
dat <- read.csv(url)
datclean <- dat |>
    rename(year = StichtagDatJahr,
           name = Vorname,
           sex = SexLang,
           births = AnzGebuWir) |>
    mutate(sex = ifelse(sex == "weiblich", "female", "male"))
```

#### Summarizing the data

We first compute the most common name by year and sex.

```{r}
summaries <- datclean |>
  group_by(sex, year) |>
  slice_max(order_by = births, n = 1) |>
  arrange(-year)
```

#### Creating a table

We can create a table summarizing our results using the `kable` function. There is either a basic version of `kabel` in the `knitr` package, or an extended version in the `kableExtra` package, which we loaded above.

```{r}
summaries |>
  kable(caption = "Most common name of newborns in the city of Zurich by year and sex")
```

Note that we could also have specified a table caption via the `tbl-cap` chunk option. Moreover, we could have specified `df-print: kable` in the YAML header to automatically apply `kable` when printing a data frame. For html reports, there is also the `df-print: paged` option that creates an interactive table. See <https://quarto.org/docs/computations/r.html#data-frames> for all possibilities.

```{r}
#| tbl-cap: "Most common name of newborns in the city of Zurich by year and sex"
## prints an interactive table if "df-print: paged" in YAML header under html options
## (only for html report but not for pdf)
summaries
```

#### Plot the data

We may want to plot the most common names between 2021 and 2024. Note that we add some code execution options for a caption (`#| fig-cap: "Most common names of Zurich newborns in 2021-2024."`) and a label (`#| label: fig-names`) so that we can cite the figure in the text.

```{r}
#| label: fig-names
#| fig-cap: "Most common names of Zurich newborns in 2021-2024."
#| message: false
datclean |>
  filter(year > 2020) |>
  group_by(sex, name) |>
  summarise(births = sum(births)) |>
  slice_max(order_by = births, n = 10) |>
  ggplot(aes(x = reorder(name, births), y = births)) +
  facet_wrap(~ sex, scales = "free") +
  geom_bar(stat = "identity") +
  labs(x = "Name", y = "Births", 
       title = "") +
  coord_flip() +
  theme_minimal()
```

We can now refer to [@fig-names] in the text using `[@fig-label]`.

### Another plot

We can also plot the occurence of a name over time. Here we choose Emma and Matteo.

```{r}
#| label: fig-names-over-time
#| fig-cap: "Popularity of the names Matteo and Emma over time."
#| message: false
datclean |>
  group_by(sex, name, year) |>
  summarise(births = sum(births)) |>
  filter(name %in% c("Emma", "Matteo")) |>
  ggplot(aes(x = year, y = births, color = name)) +
  geom_line() +
  scale_color_viridis_d() +
  labs(x = "Year", y = "Births", 
       title = "") +
  theme_minimal() +
  theme(legend.title = element_blank())
```

::: {.callout-note appearance="minimal"}
## Optional exercise

Try it with your own name!
:::

### Inline code

We can use [inline code](https://quarto.org/docs/computations/inline-code.html) to dynamically insert data summaries into text. For example,

> In 2023, the most common male name was \``` r filter(summaries, year == 2023, sex == "male")$name` `` whereas the most common female name was \``` r filter(summaries, year == 2023, sex == "female")$name` ``.

produces

> In 2023, the most common male name was `r filter(summaries, year == 2023, sex == "male")$name` whereas the most common female name was `r filter(summaries, year == 2023, sex == "female")$name`.

### References

It is also possible to include a bibliography file in BibTeX format (with .bib file extension) to automatically generate citations and a bibliography (<https://quarto.org/docs/authoring/citations.html>). This file has to be specified in the YAML header. In this demo, we have a `references.bib` file in the same directory as this .qmd file, so we specify:

``` yaml
bibliography: references.bib
```

We can then use `[@Smith2025]` (with parentheses) or `@bibkey` (without parentheses) to cite the bibliography item named Smith2025 included. For example, we recommend reading this manifesto for reproducible science [@munafo_manifesto_2017] or this nice easy-read entitled "cut with the tyranny of copy-pasting" [@perkel_cut_2022] (yes, it is about R Markdown). 

To get the best out of this functionality

-   [Connect your R studio project with a Zotero library](https://quarto.org/docs/visual-editor/technical.html#citations-from-zotero)

-   Use the Visual editor in R studio and go to insert\>citation

### Reproducibility

It is good practice to save details on software versions and computational environment when creating dynamic reports. A simple yet effective way to document your computational environment is to include the output of `sessionInfo()`.

```{r}
#| fontsize: "small"
sessionInfo()
```

## 4. Render the report

Now that you added content to your Quarto report:

-   [ ] Check that the YAML header, the text and the code chunks are well delimited and you did not forget, for example, to close any chunk with (\`\`\`).
-   [ ] Click on the button `Render` with the blue arrow on top of the editor in R studio.
-   [ ] A file with the same name as your .qmd file and the format specified in your header should have been created.
- When specifying multiple outputs, clicking on the down-arrow next to `Render` allows you to change the output format (e.g., from html to pdf).

## 5. Refine the report

Nobody knows all the options by heart, use the [Quarto documentation](https://quarto.org/docs/reference/)!

Try to explore different options for code execution:

- [ ] Hide all
- [ ] Print all code but do not execute it
- [ ] Fold all code (accessible by clicking on a button when saved as HTML)
- [ ] Hide only the code that generates the table
- [ ] **Challenge**: render the report as both HTML and PDF

# Practical tips

- Always add `sessionInfo()` or an equivalent to your report to show which packages/environment were used to generate the report
- Use relative paths
- Recompile your reports frequently to spot errors quickly
- ....

# Optional: Exploring other examples

The are many examples of dynamically generated reports online. Try exploring the links below.

- Example 1. Reproducible report in HTML using R Markdown: <https://data.snf.ch/stories/open-research-data-2023-en.html>

- Example 2. Reproducible PDF report using knitr: <https://gitlab.uzh.ch/fabio.molo/nhb-replication>

- Example 3. Website with course materials: <https://github.com/avehtari/BDA_course_Aalto/tree/master>

- Example 4. Reproducible slides with Quarto: <https://github.com/kjhealy/flipbookr-quarto/tree/main?tab=readme-ov-file>.

Try to reflect about the following questions:

-   [ ] Can you easily find the repository with the source files?
-   [ ] Which dynamic reporting tools are used?
-   [ ] Can you access the outputs easily?
-   [ ] Do they use some advanced features for dynamic reporting or computational reproducibility?
-   [ ] What would be the workflow to reproduce these reports?

# References
